---
title: "Hertie Data Club - Session 1"
author: "Benjamin Gaiser"
date: "02 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction to R

#### R is an object-oriented programming language and environment for statistical computing and graphics.

#### R provides a wide variety of statistical and graphical techniques and is highly extensible (R-packages).

#### It is great with data management and for free!

#### R is the language, and RStudio is the most popular integrated developer environment (IDE) that facilitates using R.

#### Public data is increasingly accesible and open and there are packages to directly download into your R environment.

  - e.g. World Development Indicator, GovData Germany, data.gov.uk, New York City, etc.

#### There is a huge community that can help you with problems.

  - If you ever have an error (and you will), Stackoverflow and R-Blogger will have the answer
  - Members of the community write R packages that implement state-of-the-art statistical methods or just simply make your life a lot easier

\pagebreak

### How does R look like?

#### R is divided in four quadrants

- The top left quadrant is the editor
- The bottom left quadrant is the console
- The top right quadrant is the global environment and history
- The bottom right quadrant is a multi-purpose quadrant that is mainly used for plots
- If you want, you can move them around and group the tabs however you like to fit your workflow

### How does R work?

### R is an object-oriented programme. The objects can be seen as R's nouns:

On the lowest level, data is represented in so-called atomic vectors, which form the basis of more complex data structures in R. There are four common atomic vector types:

#### numeric, also often called "double"
- represent real numbers in memory

```{r How does it work(1)}
numeric <- 10
numeric

numeric_vector <- c(1, 2.4, -3.6)
numeric_vector

class(numeric_vector)
```

#### integers (any number without decimal points; can be positive or negative)
- designated in R by adding an "L" after the number

```{r How does it work(1a)}
integer <- 10L
integer

integer_vector <- c(1L, -2L, 3L)
integer_vector

class(integer_vector)
```

#### character
- contain written characters 

```{r How does it work(2)}
character_vector <- c('Albania', 'Botswana', 'Cambodia')
character_vector
```

#### logical
- contain the logical expressions TRUE or FALSE

```{r How does it work(3)}
logical_vector <- c(TRUE, FALSE, TRUE)
logical_vector

class(logical_vector)
```

### These atomic vectors can be combined in three main ways to form more complex objects

#### Matrix
  - all vectors combined in the matrix must be equal length
  - all vectors must have the same class, otherwise *coercion* happens (we'll cover that later)
  
```{r How does it work (4)}
matrix_combined <- cbind(numeric_vector, integer_vector)
matrix_combined  # technically, the integers are now numerics, but this is not obvious here
class(matrix)  # the class of the object is a matrix
```

#### Lists
- lists can contain vectors of varying class and of varying length
```{r How does it work(6)}
test_list <- list(countries = character_vector, not_there = c(NA, NA),
                  more_numbers = 1:10)
test_list
```

#### Data Frames
- combines attributes of matrices and lists in order to easily represent statistical data
- Vectors must be of equal length, but can be of different classes 

```{r How does it work(5)}
combined_df <- data.frame(numeric_vector, character_vector, stringsAsFactors = FALSE)
combined_df
```

\pagebreak

### Statistics in R - What can it do?
```{r message=FALSE}
swiss <- as.data.frame(swiss)
library(Hmisc)
library(ggplot2)
```

#### Data Frame: Swiss Fertility and Socioeconomic Indicators (1888), R Data Set. Variables of interest:
- Independent Variable X: Catholic - % as opposed to Protestant
- Dependent Variable Y: Fertility - lg, common standardized fertility measure

#### What the data look like - Initial Descriptive Statistics
```{r message=FALSE}
summary(swiss)
```

### Hypothesis: Catholics have a higher fertility rate than Protestants.

#### First, we have a closer look at the initial descriptive statistics of our variables of interest.
```{r message=FALSE}
describe(swiss$Fertility)
describe(swiss$Catholic)
var(swiss$Fertility)
var(swiss$Catholic)
sd(swiss$Fertility)
sd(swiss$Catholic)
```

#### The variable 'Catholic' shows high variance and standard deviation for a continuous variable of between 0 and 100. Let's plot the relationship in order to see how the relationship looks.

```{r message=FALSE}
ggplot(swiss, aes(Catholic, Fertility)) + geom_point()
```


#### Adding a line of best fit, we can see that this relationship is significant, explaining 20 % of the variance (R-value). Yet, the striking thing is the clear divide between mostly Protestant and mostly Catholic cantons, except of four "outliers in the middle"

```{r message=FALSE}
ggplotRegSwiss <- function(fit){
  ggplot(swiss, aes(Catholic, Fertility)) +
    geom_point(colour = 'blue') +
    stat_smooth(method = 'lm', col = 'red', size=0.75) +
    labs(title = paste('Adj R2 = ',signif(summary(fit)$adj.r.squared, 3),
                       'Intercept =',signif(fit$coef[[1]],3 ),
                       ' Slope =',signif(fit$coef[[2]], 1),
                       ' P =',signif(summary(fit)$coef[2,4], 2)))
}
FitOfDataSwiss <- lm(Fertility ~ Catholic, data = swiss)
ggplotRegSwiss(FitOfDataSwiss)
```

\pagebreak

### Which cantons are neither mostly Protestant nor mostly Catholic?

- Step 1: Creating a factor variable with four different groups.

```{r message=FALSE}
swiss$CatholicCat <- cut(swiss$Catholic, seq(0, 100, 25))
```

- Step 2: Changing the factor variable into a character variable in order to rename the rownames.

```{r message=FALSE}
swiss$CatholicCat <- as.character(swiss$CatholicCat)
```

- Step 3: Renaming the rownames.

```{r message=FALSE}
swiss$CatholicCat[swiss$CatholicCat=='(0,25]'] <- 'Protestant'
swiss$CatholicCat[swiss$CatholicCat=='(25,50]'] <- 'Protestant to Catholic'
swiss$CatholicCat[swiss$CatholicCat=='(50,75]'] <- 'Catholic to Protestant'
swiss$CatholicCat[swiss$CatholicCat=='(75,100]'] <- 'Catholic'
```

- Step 4: Finding the cantons which are 'Protestant to Catholic' and 'Catholic to Protestant'.

```{r message=FALSE}
which(grepl('Protestant to Catholic', swiss$CatholicCat)) # 4 and 45
which(grepl('Catholic to Protestant', swiss$CatholicCat)) # 46 and 47
```

- Step 5: Searching the names.

```{r message=FALSE}
swiss[c(4,45:47),]
```

#### Moutier, V. De Geneve, Rive Droite and Rive Gauce are the only cantons where there is at least a third of the population that does not belong to the majority religion.
