---
title: "Hertie Data Club - Session 2"
author: "Benjamin Gaiser"
date: "07 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data management with R

#### Are there any questions/Have there been any issues with the exercise?

- Install the packages Hmisc and ggplot2
```{r How does it work(1)}
#install.packages("Hmisc")
#install.packages("ggplot2")
```

- Load your newly installed packages
```{r How does it work(2)}
library(Hmisc)
library(ggplot2)
```

- Create a data frame with a character, logical and numeric vector of your choice
```{r How does it work(3)}
numeric_vector <- c(1:3)
character_vector <- c("a","b","c")
logical_vector <- c(TRUE, FALSE, TRUE)
combined_df <- data.frame(numeric_vector, character_vector, logical_vector, stringsAsFactors = FALSE)
combined_df
```

- Use the help function to check out R's in-built data set "EuStockMarkets" (tip: ?object will open the Help pane)
```{r How does it work(4)}
data()
#?EuStockMarkets
```

- Load the data set "EuStockMarkets" and assign it to an object name of your choice
```{r How does it work(5)}
Stocks <- as.data.frame(EuStockMarkets)
```

- What is the mean/median/standard deviation of the four stock markets?
```{r How does it work(6)}
summary(Stocks)
sd(Stocks$DAX)
sd(Stocks$SMI)
sd(Stocks$CAC)
sd(Stocks$FTSE)
#describe(Stocks)
```

- Create a line chart for the values of the DAX (tip: use the command plot and look for the right "type" specification)
```{r How does it work(7)}
plot(Stocks$DAX, type = "l")
```

#### We will continue to use the "swiss" data set for today's Session on Data Management.

- Loading the data set and assigning it to the object "swiss"
```{r swiss(0)}
swiss <- as.data.frame(swiss)
```

- Look at the data
```{r swiss(1)}
summary(swiss)
View(swiss)
```

- Creating a new variable: High Education
```{r swiss(2)}
swiss$HighEductaion <- NA
swiss$HighEductaion[swiss$Education<12] <- 0
swiss$HighEductaion[swiss$Education>=12] <- 1
# Check if it worked out
table(swiss$HighEductaion, swiss$Education)
```

- Renaming variables
```{r swiss(3)}
names(swiss)
# column names
colnames(swiss)[colnames(swiss)=="HighEductaion"] <- "HighEducation"
names(swiss)

# row names/values
swiss$HighEducation
swiss[1,7] <- "High-level Education"
swiss$HighEducation # Have you noticed anything else?

# This is tedious and very error prone. Another method:
swiss$HighEducation[swiss$HighEducation==1] <- "High-level Education"
swiss$HighEducation
```

- Searching data
```{r swiss(4)}
summary(swiss) # Agriculture seems non linear
# Find minimum
which.min(swiss$Agriculture)
# Observation 45
swiss[45,]
# Answer: V. De Geneve

# Search for character
which(grepl("High-level Education", swiss$HighEducation))
# Look at observations
swiss[c(5,10,12,14,18,19,23,29,38,39,41,42,45,46,47),]
# Tedious. Alternative:
HighLevel <- which(grepl("High-level Education", swiss$HighEducation))
swiss[HighLevel,]
```

- Subsetting, Dropping
```{r swiss(5)}
# Only want to have a data set with the observations with High-Level Education
table(swiss$HighEducation) # 16 observations
Educated <- swiss[which(swiss$HighEducation=="High-level Education"),]
# Alternatively, we could use the subset function
Edu <- subset(swiss, HighEducation=="High-level Education")
# Drop variables
Edu$HighEducation <- NULL
summary(Edu)
# Drop observation (e.g. minimum Agriculture, i.e. observation 45)
swiss[45,]
small_swiss <- swiss[-45,]
which.min(small_swiss$Agriculture) # Observation 40
swiss[40,]
```

- Merging
```{r swiss(6)}
# Which variable does only have unique identifiers?
table(which(duplicated(swiss$Agriculture)))

# Divide into two sets
swiss_part1 <- swiss[,0:3]
swiss_part2 <- swiss[,c(2,4:7)]

names(swiss_part1)
names(swiss_part2)

# Merge by unique identifier
swiss_total <- merge(swiss_part1, swiss_part2, by="Agriculture")
```