---
title: "Hertie Data Club - Session 3"
author: "Benjamin Gaiser"
date: "14 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data visualization and analysis with R

#### Are there any questions/Have there been any issues with the exercise?

- Load packages for this Session
```{r How does it work(2)}
library(ggplot2)
library(stargazer)
```

#### We will continue to use the "swiss" data set.

- Loading the data set and assigning it to the object "swiss"
```{r swiss(0)}
swiss <- as.data.frame(swiss)
```

- Look at the data
```{r swiss(1)}
summary(swiss)
```

- Plot Fertility and Religion
```{r plot(0)}
plot(swiss$Catholic, swiss$Fertility)
```

- Create a categorial variable of low Catholocism, medium and high Catholocism
```{r cut}
# 1.Step: Cut the variable into three pices
swiss$CatholicCat <- cut(swiss$Catholic, seq(0, 100, 33.33))
# 2. Step: Changing the factor variable into a character variable for renaming the rownames
swiss$CatholicCat <- as.character(swiss$CatholicCat)
table(swiss$CatholicCat) # the four outliers are between 33 and 66 %
# 3. Step: Renaming the rownames
swiss$CatholicCat[swiss$CatholicCat=='(0,33.3]'] <- 'Low Catholicism'
swiss$CatholicCat[swiss$CatholicCat=='(33.3,66.7]'] <- 'Medium Catholicism'
swiss$CatholicCat[swiss$CatholicCat=='(66.7,100]'] <- 'High Catholicism'
# 4. Step: Finding the cantons which are 'Medium Catholicism'
MedCatholicism <- which(grepl('Medium Catholicism', swiss$CatholicCat)) # 4 and 45
MedCatholicism <- swiss[MedCatholicism,]
```

- Barplot Medium Catholicism
```{r plot(1)}
MedCatholicism

barplot <- ggplot(MedCatholicism, aes(x=Catholic, y=Fertility, fill=CatholicCat)) +
  geom_bar(stat="identity", colour="black", width=2) +
  xlab("% of people following Catholicism") + 
  ylab("Fertility rate of cantion") +
  ggtitle("Fertility rate of Medium Catholicism") +
  scale_fill_hue(name="Religious denomination") +
  theme_bw()
barplot
```

- Fancy scatterplot
```{r plot(2)}
# Remember the scatterplot from above?
plot(swiss$Catholic, swiss$Fertility)

scatter <- ggplot(swiss, aes(x=Catholic, y=Fertility))
scatter

scatter <- scatter + geom_point(aes(colour = factor(CatholicCat)))
scatter

scatter <- scatter + scale_colour_discrete(name='Catholicism')
scatter

scatter <- ggplot(swiss, aes(x=Catholic, y=Fertility, shape = factor(CatholicCat)))
scatter <- scatter + geom_point(aes(colour = factor(CatholicCat)), size = 2.5, show.legend = NA)
scatter
```

- Linear regression model
```{r regression(1),message=FALSE, results="asis"}
fit <- lm(Fertility ~ Catholic, swiss)

stargazer::stargazer(fit, title="Univariate Regression",
                     digits=2, type="html", header=FALSE)
```

- Line of best fit
```{r, message=FALSE}
scatter <- scatter + geom_smooth()
scatter

# Because the fill is Catholic Cat which is a categorical variable
bw <- ggplot(swiss, aes(x=Catholic, y=Fertility)) +
  geom_point() +
  geom_smooth()
bw

# Still not exactly what we want
lbf <- ggplot(swiss, aes(x=Catholic, y=Fertility)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)
lbf
```

